The Galactic Chemical Evolution of Carbon

### 2023-05-02 thesis/tex/main.tex
grammar edits

\usepackage{graphicx}	% Including figure files
\usepackage{booktabs}
\graphicspath{{../figures/}} 
\usepackage{isotope}
\usepackage[tableposition=above]{caption}

@@

\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=black,
    filecolor=black,
    urlcolor=black,
    citecolor=black,
    pdftitle={The Galactic Chemical Evolution of Carbon},
    pdfauthor={Daniel A Boyea},
    pdfcreator={Themself},
    pdfkeywords={astronomy,supernovae, stellar yields, chemical evolution, galaxies, carbon},
}

@@

%%%%%%%%%%%%%%% Acronyms %%%%%%%%%%%%%%%%%%
\usepackage[acronym,automake,symbols,
nogroupskip,
accsupp,
counter=section,
stylemods={longextra}
]{glossaries-extra}

\newglossary*{models}{Stellar Yield Models}
\makeglossaries

@@

\newglossarystyle{2colsymbol}{%
    \setglossarystyle{long}% base this style on the list style
    \renewenvironment{theglossary}{% Change the table type --> 3 columns
        \glsdescwidth=\dimexpr\linewidth-\dimen0-2\tabcolsep\relax
        \begin{longtable}{@{}l p{\glsdescwidth}@{}}}%
        {\end{longtable}}%
    \renewcommand*{\glossentry}[2]{%  Change the displayed items
        \glstarget{##1}{\glossentryname{##1}} %
        \glssymbol{##1}
        &  \glossentrydesc{##1}  \tabularnewline
    }
}

\newacronym
[ description ={Core collapse supernovae. Massive star explosions. {\sc ccsne} produce many elements including \gls{alpha}, \gls{fepeak}, and r and \gls{sproc}.}
]
{cc}{ccsne}{core collapse supernovae}

\newacronym
[description={Asymptotic giant branch stars. {\sc agb}s are the last phase of \glspl{lowmass}, before stars become white dwarfs. Produces C, N, and heavy s-process elements. See \ref{sec:stel_evo}.}]
{agb}{agb}{asymptotic giant branch}

\newacronym
[description={Type Ia Supernovae. Exploding white dwarfs, produces \gls{fepeak} elements and with a long delay time.}]
{ia}{sne ia}{supernovae type Ia}

@@

\newacronym
[description={Initial mass function. A function describing the mass distribution of newly formed stars. I use a \citet{kroupa01} {\sc imf}, which is described as a piecewise powerlaw function of $M$.}]
{imf}{imf}{initial mass function}

\newacronym
[description={
Apache Point Observatory Galactic Evolution Experiment. 
A large near-infrared spectroscopic survey of stars in the Milky Way. \citep{apogee17}}]
{apogee}{apogee}{Apache Point Observatory Galactic Evolution Experiment}

@@

%%%%%%%%%%%%%%%%%%% Symbols %%%%%%%%%%%%%%%%%%%%%%%%
\glsxtrnewsymbol
[description = {
    The abundance ratio between elements $A$ and $B$. $[A/B] =
    \log_{10}\left(A/B\right) - \log_{10}\left(A_{\sun}/B_{\sun}\right)$, i.e.
    $[A/B]$ is the logarithm of the ratio between A and B, scales such that
    $[A/B]=0$ for the sun. Solar abundances are as measured in
    \citet{asplund+09}.
}]
{abundratio}{\ensuremath{[A/B]}}

\glsxtrnewsymbol
[description = {The multiplicative increase in \agb\ C production. See Eq.~\ref{eq:alpha}}]
{aagb}{\ensuremath{\alpha_{\rm AGB}}}

\glsxtrnewsymbol
[description = {The multiplicative increase in \cc\ C production. See Eq.~\ref{eq:alpha}}]
{acc}{\ensuremath{\alpha_{\rm CC}}}

\glsxtrnewsymbol
[description = {The mass outflow loading factor. $\eta$ is the ratio between ejected gas from the galaxy and star formation, i.e. $\eta = \dot{M}_{\rm outflow}/\dot{M}_\star$. }]
{eta}{\ensuremath{\eta}}

\glsxtrnewsymbol
[description = {The high-mass star carbon yield metallicity dependence. See Eq.~\ref{eq:zeta}}]
{zeta}{\ensuremath{\zeta}}

\glsxtrnewsymbol
[description = {The surface density of star formation by mass.}]
{sfd}{\ensuremath{\dot{\Sigma}}}

\glsxtrnewsymbol
[description = {The mass of the sun. $M_\odot = 1.989\times10^{33}$\,g}]
{msun}{\ensuremath{M_\odot}}

\glsxtrnewsymbol
[description = {The rate of star formation (mass of new stars per time interval).}]
{sfr}{\ensuremath{\dot{M}_\star}}

\glsxtrnewsymbol
[description = {The rate of change of the mass of element $X$.}]
{mdotx}{\ensuremath{\dot{M}_{\rm X}}}

\glsxtrnewsymbol
[description = {The fractional contribution of AGB stars to the total \gls{imfave} carbon yield at solar metallicity. See Eq.~\ref{eq:f_agb}}]
{fagb}{\ensuremath{f_{\rm AGB}}}

\glsxtrnewsymbol
[description = {The galactocentric radius---distance from the galactic center in kpc.}]
{Rgal}{\ensuremath{R}}

\glsxtrnewsymbol
[description = {The per-star yield of C from \glspl{lowmass} of a given mass. In otherwords, the amount of new C an \agb\ star ejects divided by the initial mass of the star.}]
{ytcagb}{\ensuremath{\tilde{y}_{\rm C}^{\rm AGB}}}

\glsxtrnewsymbol
[description = {The \gls{imfave} yield of C from \glspl{lowmass}.}]
{ycagb}{\ensuremath{y_{\rm C}^{\rm AGB}}}

\glsxtrnewsymbol
[description = {The \gls{imfave} yield of C from \glspl{highmass}.}]
{ycc}{\ensuremath{y_{\rm C}^{\rm CC}}}

\glsxtrnewsymbol
[description = {The \gls{imfave} yield of Mg.}]
{ymg}{\ensuremath{y_{\rm Mg}^{\rm CC}}}

\glsxtrnewsymbol
[description = {The metallicity for the sun. $Z_\odot=0.014$. }]
{zsun}{\ensuremath{Z_\odot}}

\glsxtrnewsymbol
[description = {The metallicity fraction, i.e. the mass fraction of all elements which are not H or He. }]
{Z}{\ensuremath{Z}}

\glsxtrnewsymbol
[description = {The mass fraction of element $X$.}]
{Zx}{\ensuremath{Z_X}}

%%%%%%%%%%%%%%%%%%%%%% terms %%%%%%%%%%%%%%%
\newglossaryentry{metallicity}{name={metallicity},
    description={the (mass) fraction of a star or gass which is not made of either H or He. For the sun, the metallicity is $\Zo = 0.014$}
}

@@

\newglossaryentry{multizone}{name={multi-zone},
    description={the synthesis of new elements inside stars}
}

\newglossaryentry{onezone}{name={one-zone},
    description={the synthesis of new elements inside stars}
}

\newglossaryentry{alpha}{name={$\alpha$-element},
    description={Elements which are made up of $\alpha$-particles (He nuclei) throuch the triple-$\alpha$ process (see Eq. \ref{eq:triple_alpha}). Essentially light, even-numbered elements like O, Mg, and Na.}
}

@@

\newglossaryentry{rgb}{name={red giant branch},
    description={Red giant branch stars are stars which have completed hydrogen core burning and have expanded in size. See appendix \ref{sec:stel_evo}.}
}

\newglossaryentry{tdu}{name={third dredge up},
    description={Third dredge up occurs inside \gls{agb} stars. During each thermal pulse, material is {\it dredged up} from the core, changing the chemical abundances of the stellar atmosphere. (While nominally called {\it third dredge up}, there are typically several third dredge ups.). See appendix \ref{sec:stel_evo}.}
}

\newglossaryentry{fdu}{name={first dredge up},
    description={First dredge up occurs when a \gls{lowmass} enters the \gls{rgb} phase. Material from the core is brought to the surface, increasing N and decreasing C abundances. See appendix \ref{sec:stel_evo}.}
}

@@

\newglossaryentry{subgiant}{name={subgiant},
    description={A subgiant is a star which is starting to become a \gls{rgb} star.}
}

@@

\newglossaryentry{dtd}{name={delay-time distribution},
    description={The distribution in time of when an element is produced 
    after a star formation event.}
}

\newglossaryentry{massloading}{name={mass loading},
    description={The stength of outflows. See also $\gls{eta}$. }
}

@@

\newglossaryentry{lowmass}{name={low-mass star},
    description={Low-mass stars are stars with masses $\lesssim 8\,M_\odot$ which end life as a white dwarf. }
}

\newglossaryentry{highmass}{name={migh-mass star},
    description={Stars with masses $\lesssim 8\,M_\odot$, which end as a neutron star, black hole, or supernovae.}
}

\newglossaryentry{cno}{name={CNO cycle},
    description = { A proton-fusion cycle which occurs in \gls{rgb} stars 
        consisting of a chain of proton captures releasing a He nucleus ($\alpha$-particle) and energy. See Eq.~\ref{eq:cno_cycle}.
}
}
%%%%%%%%%%%%%%%%%%%%%%%%%% Models %%%%%%%%%%%%%%%%%%%%%%%
\newglossaryentry{c11}
{
    name={C11},
    type={models},
    description={the yield set from \citet{cristallo+11,cristallo+15}. The models were run at $M = $ 1.3, 1.5, 2.0, 2.5, 3.0, 4.0, 5.0, 6.0 and $Z= $ 0.0001, 0.0003, 0.001, 0.002, 0.003, 0.006, 0.008, 0.01, 0.014, 0.02. },
}
\newglossaryentry{k10}
{
    name={K10},
    type={models},
    description={the yield set from \citet{karakas10}. Includes models run at 
    $M= $ 1.0, 1.25, 1.5, 1.75, 1.9, 2.25, 2.5, 3.0, 3.5, 4.0, 4.5, 5.0, 5.5, 6.0 and $Z=$ 0.0001, 0.004, 0.008, 0.02},
}
\newglossaryentry{v13}
{
    name={V13},
    type={models},
    description={the \agb\ yield set from \citet{ventura+13,ventura+14,ventura+18,ventura+20}. Includes models with masses $m=(1.5, 2.0, 2.5, 3.0, 3.5, 4.0, 4.5, 5.0, 6.0, 6.5, 7.0)$ and with metallicities $Z=(0.0003, 0.001, 0.002, 0.004, 0.008, 0.014, 0.04)$},
}
\newglossaryentry{k16}
{
    name={K16},
    type={models},
    description={The yield set from \citet{KL16, karakas+18}. The model is sampled at 
        $M=(0.0003, 0.001, 0.002, 0.004, 0.008, 0.014, 0.04)$
        and $Z=(0.0003, 0.001, 0.002, 0.004, 0.008, 0.014, 0.04)$
    },
}

\newcommand{\cxi}{\texttt{\gls{c11}}}
\newcommand{\kx}{\texttt{\gls{k10}}}
\newcommand{\kxvi}{\texttt{\gls{k16}}}
\newcommand{\vxiii}{\texttt{\gls{v13}}}

\newcommand{\Ycc}{\gls{ycc}}
\newcommand{\Yoc}{\gls{ymg}}
\newcommand{\Ycagb}{\gls{ycagb}}
\newcommand{\sun}{\odot}
\newcommand{\Z}{Z}
\newcommand{\Zo}{Z_{\sun}}

\newcommand{\abund}[2]{\ensuremath{\[{\rm #1}/{\rm #2}\]}}
\newcommand{\proton}{\isotope[1][1]{p}}

\defcitealias{james+21}{J21}

@@

\newcommand{\citetjack}{Roberts et al.~(2023, in prep.)}
\newcommand{\citepjack}{(Roberts et al.~2023, in prep.)}
\newcommand{\citealtjack}{Robert et al.~2023, in prep.}

@@

% These dates will be filled out by the publisher
\date{\today}

@@

\begin{titlepage}
   \begin{center}
       \textbf{The Galactic Chemical Evolution of Carbon}\\
       Implications for Stellar Nucleosynthesis\\
       \vspace*{3\baselineskip}
        Undergraduate Research Thesis\\
       \vspace*{3\baselineskip}
    Presented in partial fulfillment of the requirements for graduation \textit{with research distinction in Astronomy and Astrophysics} in the College of Arts and Sciences of The Ohio State University\\
       \vspace*{3\baselineskip}
        by\\
       \vspace*{3\baselineskip}
       {Daniel Alexander Boyea}\\
       \vspace*{3\baselineskip}
       The Ohio State University\\
       April 2023\\
       \vspace*{3\baselineskip}
       Project Advisors\\
       Professor David H. Weinberg, Department of Astronomy \\
       James W. Johnson, Department of Astronomy
       \vfill
   \end{center}
\end{titlepage}

% Abstract of the paper
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}
% context
Stellar evolution models provide uncertain predictions for elemental yields due to uncertainties in stellar physics.
% aims
Here, I constrain the nucleosynthetic yields of C using multizone Galactic chemical evolution models.
% results
By matching the slopes of the \caah\ and \caafe\ relationships in a sample of APOGEE subgiant stars, I find that high-mass and low-mass stars contribute \about{80\%} and \about{20\%} of total C production respectively.
% 
To also match the normalization of the trends, I estimate the massive star C/Mg yield ratio $\Ycc/\Yoc = 1.57 + 0.59 \left(Z/Z_\odot\right)$, which explains the \caah\ trend when including the low mass C contribution.
%
Variations in the star formation history only slightly impact the low [Mg/Fe] tail of the [C/Mg]-[Mg/Fe] relation. However, most of the scatter in the \caah\ distribution can be attributed to measurement errors and radial migration. 
% 
Due to the degeneracy between the normalization of elemental yields and the strength of mass-loading in Galactic outflows, I constrain only the {\it relative} yield of C and Mg from a simple stellar population and the metallicity dependence thereof. While measurements of gas-phase C abundances are challenging, my model is broadly consistent with the [C/O]-[O/H] trend observed in compiled literature measurements.

@@

I dearly hope that the people who helped me complete this project will appreciate how much they have helped me. This year has been an incredibly challenging journey---and perhaps I am not open enough about what it has been like. I undoubtedly would not have reached this day without everyone's help.

James Johnson---your mentorship and guidance throughout this project was critical. 
You helped me learn an incredible amount about being an astronomer, and I am forever grateful for how you have guided this project, through late nights and last minute comments.  Wayne, for everything you do for everyone---you are the glue holding together our department and nobody can say thank you enough. 
David and Jennifer for helping this project move forward, by providing guidance at critical junctions and being part of my committee. Rolando, for being the non-astronomer thrown into this work, and pointing out the flaw of incomprehensibility in this work.

@@

\Gce{} aims to understand the chemical and star formation histories of galaxies. At the core of \gce{} models are stellar yields---the amount of each chemical element stars produce (or destroy). Each element is produced in different amounts by different stars, leaving traces of the Galaxy's evolutionary history. Through \textit{Galactic Archeology}, we can reconstruct the history of our galaxy by investigating clues left behind in stars.

Here, I aim to understand C enrichment---where it is produced and how its abundances evolve over time. C is unique nucleosynthetically, one of few light elements (along with N) produced in \agb{} stars \citep[e.g.][]{jennifer19, KL14}. C and N are well-studied elements since they are easy to observe in stellar spectra, even at the lowest metallicities \cite[e.g.][]{fabbian+09, nissen+14, lambert81, laird85, lambert86}. Additionally, C and N are used as age indicators in \gls{rgb} stars \citep{martig16, MG15, hasselquist19, vincenzo+21}.

Gas-phase [C/O] ratios%
%
\footnote{In this paper, I use the standard notation for chemical abundances. $[A/B] = \log_{10}\left(A/B\right) - \log_{10}\left(A_{\sun}/B_{\sun}\right)$, i.e. $[A/B]$ is the logarithm of the ratio between A and B, scales such that $[A/B]=0$ for the sun. Solar abundances are as measured in \citet{asplund+09}}%
%
---observed in very low \gls{metallicity}, high redshift \dla{}---decrease with increasing [O/H] \citep{FN15, cooke+17}. Then, [C/O] increases above $\rm [O/H] \approx -1$ (\citealt{berg+19}; see discussion in section \ref{sec:gas}).
While we know C is produced in \agb{} stars and \cc{}, we still have a limited understanding of the magnitude and metallicity-dependence of each process.

Despite their central role in \gce{} models, nucleosynthetic yields predicted by stellar evolution models are rife with uncertainties. The production rates of elements in stars are shaped by poorly understood processes, including mass loss, opacity, nuclear reaction rates, rotational mixing, and convection \citep{KL14,ventura+13, LC18}.
Changes in \gls{highmass}' explodability also introduce further variation and uncertainty in population-averaged yields \citep{emily+21}.

\Gce{} models typically use unmodified nucleosynthetic predictions and attempt to create models matching observations by varying the \sfh{} and other evolutionary parameters. Here, I introduce C yields as an additional free parameter, determining which yield prescriptions reproduce Galactic abundance trends.
\cite{james+23} examined similar \gce{} models of N, an element whose production is closely related to C. \cite{james+23} found that trends in N and O abundances can be explained by the metallicity dependence of relative N and O yields. Here, I extend their models to C and show that similar constraints on C and O relative yields can be obtained from observed abundance ratios. Additionally, I assess which yield prescriptions reproduce Galactic abundance trends while assessing the impact of \gce{} model assumptions, such as \sfh{} and outflow mass loading.

Accurate determination of stellar birth abundances for C and N is challenging. When a star enters the \gls{rgb}, material from the CNO-processed core is mixed with the envelope in first dredge up, enhancing N and depleting C \citep{iben67, vincenzo+21,KL14}. Measurements of these evolved stars'  atmospheres will no longer reflect their birth abundances.  Additionally, gas-phase measurements of C are extremely limited as C lacks strong lines in HII regions \citep{skillman+20}.

To this end, I use a sample of \gls{subgiant} stars from the \apogee{} \citep{apogee17}. According to stellar evolution theory and observations \citep{gilroy89, korn+07, lind+08, souto+18, souto19}, these stars have not yet experienced first dredge up but have well-mixed photospheres. Their photosphoric abundances should thus represent their birth composition.  In Fig.~\ref{fig:subgiants}, I plot this sample, selected by the criteria in Roberts et al. (2023, in prep; see Appendix \ref{sec:jack}). These empirical results contain information on the relative C and Mg yields, which I use later in this work.

\begin{figure}[htp]
    \centering
    \includegraphics{subgiants_mgh.pdf}
    \quad\includegraphics{subgiants_mgfe.pdf}
    \caption[APOGEE Subgiants]{The [C/Mg] ratio against [Mg/H] (top) and [Mg/Fe] (bottom) for the \citetjack~sample of \apogee{} \gls{subgiant}s. On the top, I plot high and \gls{lowalpha} stars in blue and orange, using the separation defined in Equation \ref{eq:high_alpha}. On the bottom, I color-code stars according to their [Mg/H] abundance.}
    \label{fig:subgiants}
\end{figure}

@@

Theoretical models of stellar \gls{nucleosynthesis} provide the starting point of this investigation. I focus on three primary nucleosynthetic pathways: \agb{} stars, \cc{}, and \ia. Each process has unique timescales and yields, traceable through the tools of \gce. I compare C, produced in \cc{} and \agb\ stars, to Mg and Fe as tracers of \cc{} and \ia{} respectively.

When a \ssp~forms, \cc{} are the first enrichment event. \cc{}  explode within $\lesssim 40$\,Myr, providing light elements (C, O, and Mg) and heavier elements (Fe and beyond). Metallicity-independent yields from \cc{} are the only statistically significant source of O and Mg ($\alpha$ elements). Next, low-mass stars begin to reach the end of their lives. By shedding their outer layers, \agb{} stars are important sources of C, N, and neutron capture elements.  Finally, white dwarfs explode, releasing Fe and other iron-peak elements.

>>>>

\usepackage{graphicx}	% Including figure files
\usepackage{booktabs}
\usepackage{longtable}
\graphicspath{{../figures/}} 
\usepackage{isotope}
\usepackage[tableposition=above]{caption}

@@

\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=black,
    filecolor=black,
    urlcolor=black,
    citecolor=black,
    pdftitle={The Galactic Chemical Evolution of Carbon},
    pdfauthor={Daniel A Boyea},
    pdfcreator={Themself},
    pdfkeywords={astronomy, supernovae, modeling, stellar yields, chemical evolution, galaxies, carbon},
}

@@

%%%%%%%%%%%%%%% Acronyms %%%%%%%%%%%%%%%%%%
\usepackage[acronym,automake,symbols,
nogroupskip,
accsupp,
counter=section,
stylemods={longextra}
]{glossaries-extra}
\makeglossaries

@@

\newacronym
[ description ={Core collapse supernovae. Massive star explosions. {\sc Ccsne} produce many elements including \gls{alpha}, \gls{fepeak}, and r and \gls{sproc}.}
]
{cc}{ccsne}{core collapse supernovae}

\newacronym
[description={Asymptotic giant branch stars. The {\sc agb} phase is the last phase of \glspl{lowmass}, before stars become white dwarfs. Produces C, N, and heavy \gls{sproc} elements. See \ref{sec:stel_evo}.}.]
{agb}{agb}{asymptotic giant branch}

\newacronym
[description={Type Ia Supernovae. Exploding white dwarfs. Produces \gls{fepeak} and has a long delay time.}]
{ia}{sne ia}{supernovae type Ia}

@@

\newacronym
[description={Initial mass function. A function describing the mass distribution of newly formed stars. I use a \citet{kroupa01} {\sc imf}, which is described as a piecewise power-law function of $M$.}]
{imf}{imf}{initial mass function}

\newacronym
[description={
Apache Point Observatory Galactic Evolution Experiment. 
A large near-infrared spectroscopic survey of stars in the Milky Way. \citep{apogee17}.}]
{apogee}{apogee}{Apache Point Observatory Galactic Evolution Experiment}

@@

%%%%%%%%%%%%%%%%%%%%%% terms %%%%%%%%%%%%%%%
\newglossaryentry{metallicity}{name={metallicity},
    description={the (mass) fraction of a star or gas which is not made of either H or He. For the sun, the metallicity is $\Zo = 0.014$}
}

\newglossaryentry{yield}{name={yield},
    description={The net production of a new element during a star's lifecycle divided by the star's mass (including winds and supernovae ejecta). }
}

@@

\newglossaryentry{multizone}{name={multi-zone},
    description={A chemical evolution model where a galaxy is divided into several zones, each with different stars, gas, and properties. }
}

\newglossaryentry{onezone}{name={one-zone},
    description={A chemical evolution model where the gas is all the same composition, i.e. neglecting spatial variations.}
}

\newglossaryentry{alpha}{name={$\alpha$-element},
    description={Elements which are made up of $\alpha$-particles (He nuclei) throuch the triple-$\alpha$ process (see Eq.~\ref{eq:triple_alpha}). Essentially light, even-numbered elements like O, Mg, and Na.}
}

@@

\newglossaryentry{rgb}{name={red giant branch},
    description={Red giant branch stars are stars that have completed hydrogen core burning and have expanded in size. See Appendix \ref{sec:stel_evo}.}
}

\newglossaryentry{tdu}{name={third dredge up},
    description={Third dredge up occurs inside \gls{agb} stars. During each thermal pulse, material is {\it dredged up} from the core, changing the chemical abundances of the stellar atmosphere. (While nominally called {\it third dredge up}, there are typically several third dredge ups.). See Appendix \ref{sec:stel_evo}.}
}

\newglossaryentry{fdu}{name={first dredge up},
    description={First dredge up occurs when a \gls{lowmass} enters the \gls{rgb} phase. Material from the core is brought to the surface, increasing N and decreasing C abundances. See Appendix \ref{sec:stel_evo}.}
}

@@

\newglossaryentry{subgiant}{name={subgiant},
    description={A star in the process of leaving the main sequence and becoming a \gls{rgb}.}
}

@@

\newglossaryentry{dtd}{name={delay time distribution},
    description={The distribution in time of when an element is produced 
    after a star formation event.}
}

\newglossaryentry{massloading}{name={mass loading},
    description={The strength of outflows relative to star formation. See also $\eta$. }
}

@@

\newglossaryentry{lowmass}{name={low-mass star},
    description={Stars with masses $\lesssim 8\,M_\odot$ which end life as a white dwarf. }
}

\newglossaryentry{highmass}{name={high-mass star},
    description={Stars with masses $\lesssim 8\,M_\odot$, which end as a neutron star, black hole, or supernovae.}
}

\newglossaryentry{insideout}{name={{\it insideout}},
    description={Our fiducial star formation history. The rate of star formation is highest towards the center of the galaxy and at earlier times. See Eq.~\ref{eq:inside_out}.}
}

\newglossaryentry{cno}{name={CNO cycle},
    description = { A proton-fusion cycle which occurs in \gls{rgb} stars 
        consisting of a chain of proton captures releasing a He nucleus ($\alpha$-particle) and energy. See Eq.~\ref{eq:cno_cycle}.
}
}

\defcitealias{james+21}{J21}
\newcommand{\citetjack}{Roberts et al.~(2023, in prep.)}
\newcommand{\citepjack}{(Roberts et al.~2023, in prep.)}
\newcommand{\citealtjack}{Robert et al.~2023, in prep.}

\newcommand{\cxi}{\texttt{\hyperlink{C11}{C11}}}
\newcommand{\kx}{\texttt{\hyperlink{K10}{K10}}}
\newcommand{\kxvi}{\texttt{\hyperlink{K16}{K16}}}
\newcommand{\vxiii}{\texttt{\hyperlink{V13}{V13}}}

@@

\newcommand{\Ycc}{\ensuremath{y_{\rm C}^{\rm CC}}}
\newcommand{\Yoc}{\ensuremath{y_{\rm Mg}^{\rm CC}}}
\newcommand{\Ycagb}{\ensuremath{y_{\rm C}^{\rm AGB}}}
\newcommand{\sun}{\odot}
\newcommand{\Z}{Z}
\newcommand{\Zo}{Z_{\sun}}

\newcommand{\proton}{\isotope[1][1]{p}}

@@

\date{\today}

@@

\begin{titlepage}
   \begin{center}
       \textbf{The Galactic Chemical Evolution of Carbon}\\
       Implications for Stellar Nucleosynthesis\\
       \vspace*{3\baselineskip}
        Undergraduate Research Thesis\\
       \vspace*{3\baselineskip}
    Presented in partial fulfillment of the requirements for graduation \textit{with research distinction in Astronomy and Astrophysics} in the College of Arts and Sciences of The Ohio State University\\
       \vspace*{3\baselineskip}
        by\\
       \vspace*{3\baselineskip}
       {Daniel Alexander Boyea}\\
       \vspace*{3\baselineskip}
       The Ohio State University\\
       April 2023\\
       \vspace*{3\baselineskip}
       Project Advisors\\
       James W. Johnson, Department of Astronomy \\
       Professor David H. Weinberg, Department of Astronomy 
       \vfill
   \end{center}
\end{titlepage}

% Abstract of the paper
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}
% context
Limited knowledge of stellar physics leads to uncertainties in elemental yield predictions.
% aims
As an independent approach, I constrain the nucleosynthetic yields of C using multizone Galactic chemical evolution models.
% results
By matching the median trends of \caah\ and \caafe\ in  {\sc APOGEE} subgiant stars, I find that high-mass and low-mass stars contribute \about{80\%} and \about{20\%} of total C production respectively.
% 
To match the normalization of the trends, I estimate the massive star C/Mg yield ratio $\Ycc/\Yoc = 1.57 + 0.59 \left(Z/Z_\odot\right)$,  explaining the \caah\ trend when including the low mass C contribution.
%
Variations in the star formation history only slightly impact the low [Mg/Fe] tail of the [C/Mg]-[Mg/Fe] relation. However, most of the scatter in the \caah\ distribution can be attributed to measurement errors and radial migration. 
% 
Due to the degeneracy between the normalization of elemental yields and the strength of mass-loading in Galactic outflows, I am only able to constrain the relative yields of C/Mg and its metallicity dependence. While measurements of gas-phase C abundances are challenging, my model is broadly consistent with the [C/O]-[O/H] trend observed in compiled literature measurements.

@@

I hope that the people who helped me complete this project will appreciate how much they have helped me. This year has been an incredibly challenging journey---and perhaps I am not open enough about what it has been like. I undoubtedly would not have reached this day without everyone's help.

James Johnson---your mentorship and guidance throughout this project was critical. 
You helped me learn an incredible amount about being an astronomer, and I am forever grateful for how you have guided this project, through late nights and last-minute comments.  Wayne, for everything you do for everyone---you are the glue holding together our department and nobody can say thank you enough. 
David and Jennifer for helping this project move forward, by providing guidance at critical junctions and being part of my committee. Rolando, for being the non-astronomer thrown into this work, and pointing out the flaw of incomprehensibility in this work.

@@

\Gce{} aims to understand the chemical enrichment and star formation histories of galaxies. At the core of \gce{} models are stellar \gls{yield}---the amount of each chemical element stars produce (or destroy). Each element is produced in different amounts by different stars, leaving traces of the Galaxy's evolutionary history. Through \textit{Galactic Archeology}, we can reconstruct the history of our galaxy by investigating clues left behind in stars.

Here, I aim to understand C enrichment---where it is produced and how its abundances evolve. C is unique nucleosynthetically, one of few light elements (along with N) produced in \agb{} stars \citep[e.g.][]{jennifer19, KL14}. C and N are well-studied elements since they are easy to observe in stellar spectra, even at the lowest metallicities \cite[e.g.][]{fabbian+09, nissen+14, lambert81, laird85, lambert86}. Additionally, C and N are used as age indicators in \gls{rgb} stars \citep{martig16, MG15, hasselquist19, vincenzo+21}.

Gas-phase [C/O] ratios%
%
\footnote{In this paper, I use the standard notation for chemical abundances. $[A/B] = \log_{10}\left(A/B\right) - \log_{10}\left(A_{\sun}/B_{\sun}\right)$, i.e. $[A/B]$ is the logarithm of the ratio between A and B, scaled such that $[A/B]=0$ for the sun. Solar abundances are as measured in \citet{asplund+09}.}%
%
---observed in very low \gls{metallicity}
\footnote{Astronomers are not chemists. By metallicity, I mean the (mass) fraction of any element which is not H or He, denoted by $Z$. For the sun, $Z_\odot=0.014$. }
, high redshift
\footnote{i.e. observed in the early universe.}
\dla{}---decrease with increasing [O/H] \citep{FN15, cooke+17}. Then, [C/O] increases above $\rm [O/H] \approx -1$ (\citealt{berg+19}; see discussion in section \ref{sec:gas}).
While we know C is produced in \agb{} stars and \cc{}, we still have a limited understanding of the magnitude and metallicity-dependence of each process.

Nucleosynthetic yields predicted by stellar evolution models are rife with uncertainties, despite their central role in \gce{} models, nucleosynthetic yields predicted by stellar evolution models are rife with uncertainties. The production rates of elements in stars are shaped by poorly understood processes, including mass loss, opacity, nuclear reaction rates, rotational mixing, convection, and explodability \citep{KL14,ventura+13, LC18, emily+21}.

\Gce{} models typically use unmodified nucleosynthetic predictions and attempt to create models matching observations by varying the \sfh{} and other evolutionary parameters. Here, I introduce C yields as an additional free parameter, determining which yield prescriptions reproduce Galactic abundance trends.
\cite{james+23} examined similar \gce{} models of N, an element whose production is closely related to C. They  found that trends in N and O abundances can be explained by the \gls{metallicity} dependence of relative N and O yields. Here, I extend their models to C, deriving similar constraints on C/Mg yields. I assess which yield prescriptions reproduce Galactic abundance trends while investigating the impact of \gce{} model assumptions, such as \sfh{} and outflow mass loading.

C abundances are challenging to measure. When a star enters the \gls{rgb}, material from the CNO-processed core is mixed with the envelope in first dredge up, enhancing N and depleting C \citep{iben67, vincenzo+21,KL14}. Measurements of these evolved stars'  atmospheres will no longer reflect their birth abundances.  Additionally, gas-phase measurements of C are extremely limited as C lacks strong lines in HII regions \citep{skillman+20}.

As my observational constraint, I use a sample of \gls{subgiant} stars from the \apogee{} \citep{apogee17}. According to stellar evolution theory and observations \citep{gilroy89, korn+07, lind+08, souto+18, souto19}, these stars have not yet experienced \gls{fdu} but have well-mixed envelopes. So, their surface abundances should thus represent their birth composition.  In Fig.~\ref{fig:subgiants}, I plot this sample, selected by the criteria in Roberts et al. (2023, in prep; see Appendix \ref{sec:jack}). In this sample, [C/Mg] increases with \gls{metallicity}, and [C/Mg] decreases with [Mg/Fe] at fixed [Mg/H]. Using these trends, I will develop a model of the enrichment sources and evolution of C.

\begin{figure}[htp]
    \centering
    \includegraphics{subgiants_mgh.pdf}
    \quad\includegraphics{subgiants_mgfe.pdf}
    \caption[Subgiant Abundances]{The [C/Mg] ratio against [Mg/H] (top) and [Mg/Fe] (bottom) for the \citetjack~sample of \apogee{} \gls{subgiant}s. On the top, I plot high and \gls{lowalpha} stars in blue and orange, using the separation defined in Equation \ref{eq:high_alpha} (the high and low-$\alpha$ stars are named for their high or low \gls{alpha} to Fe ratios, or in this case, Mg/Fe). On the bottom, I color-code stars according to their [Mg/H] abundance.}
    \label{fig:subgiants}
\end{figure}

@@

Theoretical models of stellar \gls{nucleosynthesis} provide the starting point of this investigation. I focus on three primary nucleosynthetic pathways: \agb{} stars, \cc{}, and \ia. Each process has unique timescales and \glspl{yield}, traceable through the tools of \gce. C is produced in both \cc\ and \agb\ stars. I also compare C to Mg and Fe, which trace \cc{} and \ia{} respectively.

When a \ssp~forms, \cc{} are the first enrichment event. \Gls{cc}  explode within $\lesssim 40$\,Myr, providing light elements (C, O, and Mg) and heavier elements (Fe and beyond). O and Mg are produced almost entirely from \cc\ with metallicity-independent yields. Next, \glspl{lowmass} stars begin to reach the end of their lives. By shedding their outer layers, \agb{} stars are important sources of C, N, and neutron capture elements.  Finally, white dwarfs explode, releasing Fe and other iron-peak elements.
